Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-06-01 VegOzone.tex
Suggested changes by Lisa

TODO: Some words to tie back to risk assment.

@@

..
This allows an understanding of which are the key aspects of modelled plant physiology that might determine the potential risk and also how the threat from ozone may change in the future. In addition, our results may hint a solution to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models and display over- or underproductivity in the Arctic \citep{GMD:Poulter2015,JAMES:Lawrence2019}.
..
Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatments ($\left<\chem{\chi_{O_3}}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Although, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamine reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species.
..

@@

Taking a presumed adaption of vegetation to local climate into consideration, we derived local parameterization for the Jarvis-type temperature and light response functions for modeling $\mathrm{POD_1}$ and biomass reduction with the $\mathrm{DO_3SE}$ model. Our key findings are that vegetation that is adapted to cold, subarctic conditions may be at higher risk of ozone damage than currently suggested by UNECE risk assessment methods as such vegetation can be physiologically active even under cold temperature conditions; the occurrence of spring ozone peaks enhance the likelihood of ozone uptake and potential damage. TODO: On the other hand, a warmer climate and hence prolonged GS would negatively affect any vegetation using the $\mathrm{POD_y}$ metric. It is essential for ozone risk assessments in the subarctic to define plant physiology and traits such as phenological states properly. As no independent ozone flux measurements are available at Svanhovd, we could not fully validate our methodology at this point. 
A comparison of $\mathrm{DO_3SE}$ model derived dry deposition velocities $V_\mathrm{d}$ for coniferous trees with comprehensive records over Pine forests from Hyytilälä, Finland \citep{BER:Keronen2003} (refer to supplementary material Figs.~S1--S3) indicate a slightly too high $V_\mathrm{d}$ in spring for our most extreme localization (subarctic). No comparable data is available for northern Fennoscandia or deciduous tress and perennial grassland. Perennial grassland and deciduous trees displayed strong susceptibility to our localization methodology (increase in frequency of maximum $V_\mathrm{d}$ throughout the GS, see supplementary material Figs.~S1--S3) but are within the range ($v\mathrm{d} = (2-9)\,\unit{mm\,s^{-1}}$ for grassland) given in the literature \citep[e.g.][]{JGR:Jacob1992,AE:Padro1996,RG:Clifton2020}. Independent flux measurements over these vegetation types in a subarctic climate are thus crucial for evaluation and validation.

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone injuries on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher ozone damage risk under warmer climate conditions for these specific plant species in a subarctic climate. However, visible injuries are not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007}. Also, our $\mathrm{DO_3SE}$ modeling results are in contrast to these observations at Svanhovd. Any visible injuries caused by peak ozone flux through stomata or a specific dependency of plant sensibility to ozone on phenology stage are not accounted for in the $\mathrm{DO_3SE}$ model. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible injuries explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. Forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed visible injuries in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

@@

We found that the difference in biomass reduction compared to Boreal parmeterizations \citep{ICP:MappingManual2017} could amount to $55\,\unit{\%}$ in deciduous trees and $29-43\,\unit{\%}$ in perennial grassland. But the biomass reduction functions were established based on studies from less extreme climates, hence, these results ought to be treated with care. In particular, CLs for deciduous and coniferous trees were breached in both years and for all parameterizations. For perennial grassland, the CL was not breached if a localized GS was assumed, but the defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constraints on productivity leading to more severe economic consequences.

Beyond the risk assessment of ozone-induced damage, our localized Jarvis-type temperature response functions, once verified by observations, can have important implications on land-surface modeling still using such parameterizations in global models. Problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. A verification of our proposed methodology by independent flux and dry deposition measurements over birch forest and perennial grassland would be necessary to establish a climate data based localization of stomatal conductance parameterizations.

>>>>

\subsection{Implications on ozone damage risk}
\label{subsec:clim_risk}

As we have shown, 2019 was a rather normal year compared to 2018 with respect to present-day climate and ozone concentrations. In conjunction with the $\mathrm{DO_3SE}$ modeling results (Section~\ref{subsec:do3se_results}), this indicates that risk of ozone-induced damage in subarctic vegetation has been underestimated in assessments using the Boreal MM parameterizations. In future scenarios in which persistent heatwaves like in 2018 become more common, the risk for ozone-induced damage is probably reduced for vegetation adapted to cold growing conditions. A longer growing season promoted by climate change without weather extremes like in 2019 is likely to increase this risk.

@@

@@

Taking a presumed adaption of vegetation to local climate into consideration, we derived local parameterization for the Jarvis-type temperature and light response functions for modeling $\mathrm{POD_1}$ and biomass reduction with the $\mathrm{DO_3SE}$ model. Our key findings are that vegetation that is adapted to cold, subarctic conditions may be at higher risk of ozone damage than currently suggested by UNECE risk assessment methods as such vegetation can be physiologically active even under cold temperature conditions; the occurrence of spring ozone peaks enhance the likelihood of ozone uptake and potential damage. On the other hand, a warmer climate and longer GS increase $\mathrm{POD_y}$ and hence risk. It is essential for ozone risk assessments in the subarctic to define plant physiology and traits such as phenological states properly. As no independent ozone flux measurements are available at Svanhovd, we could not fully validate our methodology at this point. 
Dry deposition velocities $V_\mathrm{d}$ for coniferous trees from comprehensive records over Pine forests from Hyytilälä, Finland \citep{BER:Keronen2003} yield summer maximum $V_\mathrm{d}\sim 11\,\unit{mm\,s^{-1}}$. The $\mathrm{DO_3SE}$ model derived summer maximum $V_\mathrm{d}\sim 9\,\unit{mm\,s^{-1}}$ with minimal differences between the parameterizations. The highest $V_\mathrm{d}$ is found in summer 2019 using the cold paramterization. The maximum observed spring $V_\mathrm{d}\sim 8\,\unit{mm\,s^{-1}}$ is slightly smaller compared to $V_\mathrm{d}\sim 10\,\unit{mm\,s^{-1}}$ from the $\mathrm{DO_3SE}$ model. No comparable data is available for northern Fennoscandia or deciduous tress and perennial grassland. Perennial grassland and deciduous trees displayed strong susceptibility to our localization methodology (increase in frequency of maximum $V_\mathrm{d}$ throughout the GS, see supplementary material Figs.~S1--S3) but are within the range ($v\mathrm{d} = (2-9)\,\unit{mm\,s^{-1}}$ for grassland) given in the literature \citep[e.g.][]{JGR:Jacob1992,AE:Padro1996,RG:Clifton2020}. Independent flux measurements over these vegetation types in a subarctic climate are thus crucial for evaluation and validation.

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone injuries on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher risk of ozone damage under warmer climate conditions for these specific plant species in a subarctic climate. However, visible injuries are not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007} and the $\mathrm{DO_3SE}$ model does not estimate visible injury which is often related to short periods of very high ozone uptake. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible injuries explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. The \citet{Amb:Manninen2009} results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatments ($\left<\chem{\chi_{O_3}}\right> = 36-54\,\unit{ppb}$). Although, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamine reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species. Forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed visible injuries in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function relating ozone susceptibility to leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions. Such a damage function could be different from $f_\mathrm{phen}$.

@@

We found that the difference in biomass reduction compared to Boreal parmeterizations \citep{ICP:MappingManual2017} could amount to $55\,\unit{\%}$ in deciduous trees and $29-43\,\unit{\%}$ in perennial grassland. But the biomass reduction functions were established based on studies from less extreme climates, hence, these results ought to be treated with care. In particular, CLs for deciduous and coniferous trees were breached in both years and for all parameterizations. For perennial grassland, the CL was not breached, but the defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constraints on productivity leading to more severe economic consequences.

Beyond the risk assessment of ozone-induced damage, our climate derived local parameterization Jarvis-type temperature response functions, once verified by observations, can have important implications on land-surface modeling still using only Boreal parameterizations in global models. Problems in estimating the productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. A verification of our proposed methodology by independent flux and dry deposition measurements over birch forest and perennial grassland would be necessary to establish a climate data based localization of stomatal conductance parameterizations.
